---
title: "Sleep stress"
author: "Riya"
date: "2023-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r}
library(arules)
library(dplyr)
library(tidyverse)
```

### Reading and cleaning dataset

```{r}
dfs<-read.csv(file="https://raw.githubusercontent.com/riyaeliza123/sleepstress/main/data/SLEEP_QUALITY.csv")
#View(dfs)

#Giving new column names
colnames(dfs) <- c('Timestamp','Gender','Age','Major','Diagnosed?','Treatment','Bedtime','Fall_asleep','Wakeup_time','Sleep_time','Sleep.30min','Sleep.wakeUpMiddle','Sleep.bathroom','Sleep.snore','Sleep.cold','Sleep.hot','Sleep.badDreams','Sleep.pain','Stress.assignment','Stress.exams','Stress.ketchup','Stress.newPerson','Stress.help','Stress.confronting','Stress.noHelp')

#copying dfs into temp to manipulate dfs
temp <- dfs 

#Converting categorical columns to numerical categories

levels <- c('Not during the past month','Less than once a week',
                                'Once or twice a week','Three or more times a week')
labels <- c(1,2,3,4)

dfs$Sleep.30min=factor(dfs$Sleep.30min, levels = levels, labels = labels)

dfs$Sleep.wakeUpMiddle=factor(dfs$Sleep.wakeUpMiddle, levels = levels, labels = labels)

dfs$Sleep.bathroom=factor(dfs$Sleep.bathroom, levels = levels, labels = labels)

dfs$Sleep.snore=factor(dfs$Sleep.snore, levels = levels, labels = labels)

dfs$Sleep.cold=factor(dfs$Sleep.cold, levels = levels, labels = labels)

dfs$Sleep.hot=factor(dfs$Sleep.hot, levels = levels, labels = labels)

dfs$Sleep.badDreams=factor(dfs$Sleep.badDreams, levels = levels, labels = labels)

dfs$Sleep.pain=factor(dfs$Sleep.pain, levels = levels, labels = labels)

dfs$Fall_asleep = factor(dfs$Fall_asleep, 
                         levels=c('under 30 minutes','30 minutes','1 hour','1.5 hours', 
                                  '2 hours', 'More time than that'), 
                         labels=c(1,2,3,4,5,6))
dfs$Sleep_time = factor(dfs$Sleep_time, 
                        levels=c('less than 4 hours of sleep','4-6 hours',
                                 '7-8 hours','more than 8 hours'), 
                        labels=c(1,2,3,4))
```

```{r}
View(dfs)
```

### Creating df.sleep containing all sleep related columns

```{r}
df.sleep<-data.frame(dfs[8],dfs[10:18])
#View(df.sleep)
df.sleep_num <- lapply(df.sleep,as.numeric) #convert to numeric list

df.sleep <- data.frame(df.sleep_num) #convert back to dataframe

df.sleep$INDEX_sleep <- rowSums(df.sleep, na.rm=TRUE)
View(df.sleep)
```

### Normalizing INDEX and creating a new column with the normalized values

```{r}
# Define a function to normalize a column to a specified range
normalize_to_range <- function(column, min_value, max_value, new_min, new_max) {
  # Normalize the values to the new range
  normalized_values <- ((column - min_value) / (max_value - min_value)) * (new_max - new_min) + new_min
  
  # Round the normalized values to the nearest integer
  normalized_values <- round(normalized_values)
  
  # Ensure the values are within the specified range (1 to 5)
  normalized_values <- pmax(pmin(normalized_values, new_max), new_min)
  
  return(normalized_values)
}

# Specify the range for the new values (1 to 5)
new_min <- 1
new_max <- 5

# Get the minimum and maximum values of the column you want to normalize
min_value <- min(df.sleep$INDEX_sleep)
max_value <- max(df.sleep$INDEX_sleep)

# Normalize the specified column
df.sleep$normINDEX_sleep <- normalize_to_range(df.sleep$INDEX_sleep, min_value, max_value, new_min, new_max)

View(df.sleep)

```

### Encoding data for df.stress

```{r}
levels = c('not stressed','mild','moderate','severe','very severe')
labels = c(1, 2, 3, 4, 5)

dfs$Stress.assignment=factor(dfs$Stress.assignment, levels = levels, labels = labels)

dfs$Stress.exams=factor(dfs$Stress.exams, levels = levels, labels = labels)

dfs$Stress.ketchup=factor(dfs$Stress.ketchup, levels = levels, labels = labels)

dfs$Stress.newPerson=factor(dfs$Stress.newPerson, levels = levels, labels = labels)

dfs$Stress.help=factor(dfs$Stress.help, levels = levels, labels = labels)

dfs$Stress.confronting=factor(dfs$Stress.confronting, levels = levels, labels = labels)

dfs$Stress.noHelp=factor(dfs$Stress.noHelp, levels = levels, labels = labels)

```

### Creating the stress dataframe

```{r}
df.stress <- data.frame(dfs[19:25])

#convert to numeric list
df.stress_num <- lapply(df.stress,as.numeric) 

#convert back to dataframe
df.stress <- data.frame(df.stress_num) 

df.stress$INDEX_stress <- rowSums(df.stress, na.rm=TRUE)
#View(df.stress)
```

### Normalizing stress INDEX

```{r}
min_value <- min(df.stress$INDEX_stress)
max_value <- max(df.stress$INDEX_stress)

# Normalize the specified column
df.stress$normINDEX_stress <- normalize_to_range(df.stress$INDEX_stress, min_value, max_value, 
                                                new_min, new_max)

#View(df.stress)
```

### Merging 2 dataframes to find total index, dataframe is final_index

```{r}
final_index<-data.frame(df.sleep$normINDEX_sleep, df.stress$normINDEX_stress)
final_index$TOTAL_INDEX<- rowSums(final_index, na.rm=TRUE)
```

### Getting final dataframe to derive conclusions

```{r}
final<-data.frame(dfs[2:4],final_index,dfs[5:6])
#colnames(final)

colnames(final) <- c('Gender','Age','Major','Sleep_index','Stress_index','Total','Diagnosed','Treatment')
#View(final)

# Merge sleep_index, stress_index, total_index ---> dataframe called complete
complete<-data.frame(temp,final[4:6])
colnames(complete)[colnames(complete) == "Diagnosed."] <- "Diagnosed"
```

# QUESTIONS FOR CONCLUSIONS:

#### 1. For people who have been professionally diagnosed with a mental illness, what is their stress and sleep pattern? Who else (non-diagnosed) has same pattern as diagnosed people, hence, may indicate possibility of mental illness

```{r}
 diag_yes <- filter(complete, Diagnosed == 'Yes')
 #View(diag_yes)
 #avg total score=?
 mean = mean(diag_yes$Total) #6.7778
 
 paste0("The mean sleep_stress index of people diagnosed of a mental illness is ", round(mean, 2))
 
 #take a list of non-diagnosed above score 6
 non_diag_count <- count(filter(complete, Diagnosed == 'No'))
 notdiag_high <- filter(complete, Diagnosed == 'No', Total > mean)
 count_notdiag_high = count(notdiag_high)
 
 paste0(count_notdiag_high, " out of ", non_diag_count, 
        " people have the same stress indices as diagnosed individuals.")
 paste0("That is, ", round(count_notdiag_high*100/non_diag_count, 2), "% of non diagnosed people show signs of mental distress.")
```

### 2. Analyse people with irregular sleep patterns (sleep between 1am-3am and sleep for upto 6 hrs)

```{r}
latesleepers<-filter(complete, Bedtime=='1am-3am', 
                     Sleep_time %in% c("less than 4 hours of sleep", "4-6 hours"))
 #View(latesleepers)
 
 total_df = count(complete)
 
 count_latesleep = count(latesleepers)
 paste0(round(count_latesleep*100/total_df, 2), "% of people have less sleep.")
 
 
 #check how many have total index above 6, meaning show signs of mental distress
 
 count_above6 = count(filter(latesleepers,Total>6)) 
 count_above6_yes = count(filter(latesleepers,Total>6, Diagnosed == "Yes")) 
 
 #Conclusion
 paste0(count_above6," out of ", count_latesleep, 
        " show signs of mental distress. i.e; ", 
        round(count_above6*100/count_latesleep), 
        "% of late sleepers show signs of mental distress, out of which, ", 
        round(count_above6_yes*100/count_latesleep), 
        "% have been diagnosed.")
 
```

### 3.  The number of people diagnosed is not equal to the number of people treated. Why? Patterns?

```{r}
 treatment_yes <- filter(complete, Treatment=='Yes')
 treatment_yes
 
 treatment_diag <- filter(treatment_yes, Diagnosed == 'Yes')
 
 paste0("There are ", count(treatment_yes), 
        " individuals who seeked treatmeant but only ", 
        count(treatment_diag), 
        " of them were diagnosed with a mental illness")
```

There is very limited data about people who got treatment, but if we ran a survey just for them, asking more details like time and circumstance of diagnosis, we can draw better conclusions

### 4. Is there a pattern in people who take more than 2 hrs to fall asleep? What is the demographic like?

```{r}
moresleep <- filter(complete, Fall_asleep=='More time than that')
moresleep
```

The data is limited, but the question has scope and must be explored in a bigger dataset.

## Visualization

### A. Visualization of the statistics calculated

#### 1.  Diagnosed v/s undiagnosed individuals

```{r}
count_data <- data.frame(
  Category = c("Diagnosed", "Un-diagnosed"),
  Count = c(z, y)
)

# Create a pie chart
colors <- c("lightgreen", "lightblue")
pie_labels <- paste(count_data$Category, ": ", count_data$Count) 

pie(count_data$Count,
    labels = pie_labels,
    col = colors,
    main = "Diagnosed vs. Undiagnosed in the Total Strength",
)

legend("topright", count_data$Category, fill = colors, title = "Category")

```

#### 2. Among undiagnosed people, what percentage show signs of mental illness?

```{r}

# Define the data
total <- y
possible_ill <- as.integer(count(notdiag_high))
fit <- as.integer(total - possible_ill)
x <- c(fit, possible_ill)
labels <- c("Low", "High")
colors <- c("lightgreen", "lightblue")

# Create the pie chart
pie(x, labels = labels, col = colors, main = "Potential for Mental Illness Among Non-Diagnosed Individuals")

# Add a legend
legend("topright", labels, fill = colors, title = "Potential")

```
